---
title: "Get Sequences"
author: "Cesaré Ovando-Vázquez"
date: '`r format(Sys.Date(), "%d/%B/%Y")`'
output: html_document
editor_options: 
  chunk_output_type: console
---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# library(goseq)
library(rtracklayer)
library(GenomicFeatures)
library(reshape2)
library(Biostrings)
library(GO.db)
library(gplots)
library(RColorBrewer)

rm(list = ls())
gc()

# Paths
# inPath <- "../08.edger/"
outPath <- "GetSeqs"

dir.create(outPath)
```


We extract pre-mRNA and mRNA sequences from genome fasta using StringTie gtf results. 

Our objetive is to get pre-mRNA and mRNA sequences. mRNA to generate the transcriptome reference to **kallisto** and pre-mRNA to be used in further physics (folding energy, entropy, stability, etc.) calculations.

We get exons regions per transcript with unique identifier, then we export these granges and use **bedtools** to get sequences.


## Get genomic regions

We import StringTie results annotation, GTF file.
```{r}
# Get gtf
gff <- import("../Genome/grs.gff")
genes <- gff[gff$type == "gene"]
trx <- gff[gff$type == "mRNA"]
exons <- gff[gff$type == "exon"]
```


Get Gene to transcript map. Mandatory to annotate transcript with Trinotate

```{r}
# Get Gene to Transcript map (two columns with no header)
gene_trans_mapTab <- data.frame(Gene = sub("gene:", "", unlist(trx$Parent)), Transcript = sub("mRNA:", "", trx$ID), stringsAsFactors = FALSE)

write.table(x = gene_trans_mapTab, file = paste0(outPath, "/gene_trans_map"), row.names = FALSE, col.names = FALSE, quote = FALSE, sep = "\t")
```


Get genes from StrongTie to reference genes mapping

```{r}
# Mapping between gene's IDs
mbo <- mergeByOverlaps(subject = genes, query = trx)

# Get map
StringTie2Gene <- sapply(split(mbo$Name, f = mbo$Name.1), function(x) paste(x, collapse = "|"))

# Get table
mapTab <- data.frame(StrinTie = names(StringTie2Gene), gene_name = StringTie2Gene, stringsAsFactors = FALSE)

write.table(x = mapTab, file = paste0(outPath, "/StringTie2Gene_map.tab"), sep = "\t", col.names = TRUE, quote = FALSE)
```



We set these lists to a list to process 
```{r}
# Set lists to list
grPerList <- list(pre_mrna = trx, mrna = exons)
```

Export granges as gff 
```{r}
# Export granges
for (reg in names(grPerList)) {
  
  # Get granges 
  grs <- grPerList[[reg]]
  
  # Set type
  grs$type <- paste0(grs$Parent ,"_", grs$ID)
 grs$type <- paste0(grs$type, "-", ifelse(strand(grs) == "+", "pos", "neg"))
  
  # Export  
  export(object = grs, paste0(outPath, "/", reg, ".gff"))
}
```



## Get sequence from genome fasta

Once exported , we use **bedtools** to ge sequences on the system
```{r}
# Use bedtools to get sequences
for (reg in names(grPerList)) {
  
  # Run bedtools
  system(paste0("bedtools getfasta -s -name -fi ../Genome/genome.fasta -bed GetSeqs/", reg, 
                ".gff -fo GetSeqs/", reg, ".fa"))
}
```



## Import sequences and get 

We import sequences and save it. If exons, we concat to get mRNA. Transcripts are pre-mRNA.
```{r}
# Read sequences
for (reg in names(grPerList)) {
  
  # Read exon sequences
  regSeq <- readDNAStringSet(paste0("GetSeqs/", reg, ".fa"))
  
  # Split exons by transcript, if reg == exons
  if (reg == "mrna") {
    
    # Split
    exonsSeqList <- split(regSeq, f = gsub("mRNA:|\\_.+", "", names(regSeq)))
    
    # Concat exons seqs 
    mRNASeqList <- lapply(exonsSeqList, function(seq) {
      if ((grepl("neg", names(seq)))[1]) {
        mrnaSeq <- paste(rev(as.character(seq)), collapse = "")
      } else {
        mrnaSeq <- paste((as.character(seq)), collapse = "")
      }
      return(mrnaSeq)
    })
        names(mRNASeqList) <- sub("-.+$", "", names(mRNASeqList))
   
    regSeq <- DNAStringSet(x = unlist(mRNASeqList), use.names = TRUE)
  }
  
  # Get insight about sequences
  w <- width(regSeq)
  w <- ifelse(w > 15e3, 15e3, w)
  
  # Histogram
  hist(w, breaks = 64, main = paste0(reg, "\nMean = ", mean(w)))
  abline(v = mean(w))
  grid()
  
  # Ecdf
  plot(ecdf(w), main = paste0(reg, "\nMedian = ", median(w)))
  abline(v = median(w), h = 0.5)
  grid()
  
  # Write XStringSet
  print(paste0("Writing ", length(regSeq), " sequences"))
  writeXStringSet(regSeq, paste0("GetSeqs/", reg, "_Seqs.fa"))
}
# # RNAfold
# trxIds <- names(trxSeq)
# for (trxId in trxIds) {
#   system(paste("RNAfold", as.character(trxSeq[names(trxSeq) == trxId])))
# }
```

