---
title: "GetCounts"
author: "Cesaré Ovando-Vázquez"
date: '`r format(Sys.Date(), "%d/%B/%Y")`'
output:
  html_document:
    theme: cerulean
    highlight: haddock
editor_options: 
  chunk_output_type: inline
---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(reshape2)
library(factoextra)
library(RColorBrewer)
# library(scatterplot3d)

rm(list = ls())
gc()

# Paths
inPath <- "../01.01.kallisto/"
outPath <- "GetCounts"

# Make paths
dir.create(outPath, showWarnings = FALSE)

# Patterns
inPatt <- ".tsv$"
# outPatt <- inPatt
```


# List Files

Get kallisto results (.tsv files), add library information.

```{r}
# List featureCounts results
print("List Files")
fcFiles <- list.files(path = inPath, pattern = inPatt, full.names = TRUE, recursive = TRUE)

# Concat files
# countList <- list()
cmd <- "cat ../01.01.kallisto/kallisto/Arcanum_T0h_P1/abundance.tsv | cut -f1"
trxids <- system(cmd, intern = TRUE)
countTab <- data.frame(row.names = trxids)
# countTab <- NULL
for (fcFile in fcFiles) {
  
  print(fcFile)
  
  # Load fcFile
  tab <- read.table(fcFile, header = TRUE, sep = "\t", quote = "")
  head(tab)
  
  # Re-name columns
  colnames(tab) <- sub("^.+\\.bam$", "counts", colnames(tab))
  
  # Keep columns of interest
  tab <- tab[, c("target_id", "est_counts")]
  
  # Add library name
  # tab$lib <- basename(dirname(fcFile))
  
  # Add tab to list
  # countList[[fcFile]] <- tab
  countTab <- cbind(countTab, tab[row.names(countTab), "est_counts"])
}
# row.names(countTab) <- tab$target_id
# colnames(countTab) <- basename(dirname(fcFiles))
head(countTab)
```

Get count table
```{r}
# Concatenate
# countTab <- do.call(rbind, countList)
# countTab$counts <- countTab$est_counts
```

# Summary countTab

Get total counts per library.
```{r}
# Total library counts
# totalCounts <- sum(countTab$counts)/1e6
# totalCountsLib <- aggregate(countTab$counts, by = list(countTab$lib), sum) 

# Plot totalCountsLib
par(omi = c(1, 0, 0, 0))
# barplot(totalCountsLib$x, border = FALSE, col = "dodgerblue1", names.arg = totalCountsLib$Group.1, las = 3)
grid()
```

# Get count matrix

We get and save count matrix to be used by edgeR. Then, we CPM-normalize libraries.

```{r}
# Get counts per gene id and library
# totalCountsTrxLib <- aggregate(counts ~ lib + target_id, data = countTab, sum) 

# Get matrix
countMat <- acast(data = countTab, formula = target_id ~ lib, fun.aggregate = sum, value.var = "counts")
row.names(countMat) <- sub("^.*gene\\:", "", row.names(countMat))
colnames(countMat) <- sub("Arcanum", "S._arc", colnames(countMat))
colnames(countMat) <- sub("Slyc", "S._lyc", colnames(countMat))
colnames(countMat) <- sub("T", "", colnames(countMat))
head(countMat)

# Write counts table
countMat2save <- data.frame(GeneID = row.names(countMat), round(countMat))
write.table(countMat2save, file = paste0(outPath, "/counts.tab"), sep = "\t", quote = FALSE, row.names = FALSE, col.names = TRUE)

# Get CPMs
cpmMat <- apply(countMat, 2, function(x) x/sum(x)*1e6)

# Write cpm table
cpmMat2save <- data.frame(GeneID = row.names(cpmMat), round(cpmMat))
write.table(cpmMat2save, file = paste0(outPath, "/cpm.tab"), sep = "\t", quote = FALSE, row.names = FALSE, col.names = TRUE)
```

# Explore data

Visualize library variabilty using boxplots.

```{r fig.height = 6, fig.width = 6, fig.align = "center"}
# Plot
# pdf(paste0(outPath, "/plots.pdf"))

# Count boxplot
par(omi = c(1, 0, 0, 0), mfrow = c(1, 1))
boxplot(log10(countMat+1), las = 3, cex = 0.5, 
        col = ifelse(grepl("Arc", colnames(countMat)), 
                     rgb(1, 0, 0, 0.5, 1), rgb(0, 0, 1, 0.5, 1)), 
        main = "Counts")

# CPM boxplot
par(omi = c(1, 0, 0, 0), mfrow = c(1, 1))
boxplot(log10(cpmMat+1), las = 3, cex = 0.5, 
        col = ifelse(grepl("Arc", colnames(countMat)), 
                     rgb(1, 0, 0, 0.5, 1), rgb(0, 0, 1, 0.5, 1)),
        main = "CPM")
```

Explore between libraries variability. We show three examples.

```{r fig.height = 6, fig.width = 6, fig.align = "center"}
# Scatter plot genes
col <- rgb(0.1, 0.1, 0.1, 0.25)
plot(log10(cpmMat[, c(1, 2)]+1), col = rgb(0.1, 0.1, 0.1, 0.3), pch = 20, main = "Scatter plot log-log")
grid()
abline(a = 0, b = 1)

plot(log10(cpmMat[, c(1, 4)]+1), col = rgb(0.1, 0.1, 0.1, 0.3), pch = 20, main = "Scatter plot log-log")
grid()
abline(a = 0, b = 1)

plot(log10(cpmMat[, c(1, ncol(cpmMat))]+1), col = rgb(0.1, 0.1, 0.1, 0.3), pch = 20, main = "Scatter plot log-log")
grid()
abline(a = 0, b = 1)
```

Finally, we use Principal Component Analysis (PCA) to:
. Observe quality of data, "correct" clustering between samples
. Identify potential latent variables

```{r  fig.height = 6, fig.width = 6, fig.align = "center"}
# Filter out genes with 0 counts
cpmMatFiltered <- cpmMat[rowSums(cpmMat >= 4) >= 12,]

# PCA bith species
pcaCpmMat <- prcomp(t(cpmMatFiltered), scale. = TRUE)

# Plot variabity per PC
barplot(pcaCpmMat$sdev**2, border = FALSE, cex.names = 0.8, las = 3,
        names.arg = paste0(colnames(pcaCpmMat$rotation), " - ", 
                           round(pcaCpmMat$sdev**2/sum(pcaCpmMat$sdev**2)*100, 1), "%"), 
        ylab = "Variances")
grid()

# PCA plot all samples plot
fviz_pca_ind(pcaCpmMat, habillage = sub("\\_P[0-9]+", "", colnames(cpmMatFiltered)), addEllipses = TRUE,
              title = "", repel = TRUE)

# PCA per species
lycMat <- cpmMatFiltered[, grepl("S._lyc", colnames(cpmMatFiltered))]
pcaLyc <- prcomp(t(lycMat), scale. = TRUE)
arcMat <- cpmMatFiltered[, grepl("S._arc", colnames(cpmMatFiltered))]
pcaArc <- prcomp(t(arcMat), scale. = TRUE)

# PCA plot per species
fviz_pca_ind(pcaLyc, habillage = sub("\\_P[0-9]+", "", colnames(lycMat)), addEllipses = TRUE)
fviz_pca_ind(pcaArc, habillage = sub("\\_P[0-9]+", "", colnames(arcMat)), addEllipses = TRUE)

#
# labels <- unique(sub("\\_P[0-9]+", "", colnames(geneMat)))
# cols <-  brewer.pal(length(labels), "Set1")
# names(cols) <- labels

# #
# par(mfrow(c(1, 1)))
# scatterplot3d(pcaGeneMat$x[, 1:3], color = cols[sub("\\_P[0-9]+", "", colnames(geneMat))], pch = 20, cex.symbols = 3, type = "h")
# legend("bottomright", legend = labels, fill = cols, border = FALSE, bty = "n")
# 
# par(mfrow(c(1, 2)))
# scatterplot3d(pcaLyc$x[, 1:3], color = cols[sub("\\_P[0-9]+", "", colnames(lycTab))], pch = 20, cex.symbols = 3, type = "h")
# legend("bottomright", legend = labels[grepl("Lyc", labels)], fill = cols[labels[grepl("Lyc", labels)]], border = FALSE, bty = "n")
# scatterplot3d(pcaArc$x[, 1:3], color = cols[sub("\\_P[0-9]+", "", colnames(arcTab))], pch = 20, cex.symbols = 3, type = "h")
# legend("bottomright", legend = labels[grepl("Arc", labels)], fill = cols[labels[grepl("Arc", labels)]], border = FALSE, bty = "n")

```

### Session information
```{r}
sessionInfo()
```




